Determinism, Complexity, and Predictability in Computer Performance

### 2014-04-12 concl.tex
first draft of conclusion

It's obvious that no forecast method can work if a time series has no
predictive structure, but it's hard to define or measure predictive
structure.  In this paper, we have argued that redundancy---insert
short definition here---is a useful definition of predictive structure
in a time series, and that weighted permutation entropy is an
effective way to estimate how much of that structure is present in a
given real-valued, scalar time series.  Even if predictive structure
exists, a given forecast method may not work---simply because it
cannot capture and/or exploit the structure that is present (e.g., a
linear model of a nonlinear process).  This fairly obvious point can
be a serious practical challenge.  When a model isn't working, the
practitioner doesn't know whether it's because there's no predictive
structure (i.e., no model will work) or whether the model simply isn't
good enough.  Fortunately, weighted permutation entropy allows us to
make this distinction for noisy real-valued time series.

In particular, we wish to determine---without knowing anything about
the generating process---whether a time series is too complex to
successfully predict.  We have argued that
redundancy~\cite{crutchfield2003} is a useful definition of the
inherent predictability of a time series.  To operationalize that
definition, we use an approximation of the Kolmogorov-Sinai
entropy~\cite{lind95}, estimated using a weighted version of the
permutation entropy of~\cite{bandt2002per}.  This WPE technique---an
ordinal calculation of forward information transfer in a time
series---is ideal for our purposes because it works with real-valued
data and is known to converge to the true entropy value.  Using a
variety of forecast models and more than a hundred time-series data
sets from a computer-performance experiment, we have shown that
prediction accuracy is indeed correlated with weighted permutation
entropy: the higher the WPE, in general, the higher the prediction
error.  There are some exceptions: summarize \& discuss here.

@@

An important practical corollary to this empirical correlation of
predictability and WPE is a strategy for assessing appropriateness of
forecast methods.  If the forecast produced by a particular method is
poor but the time series contains a significant amount of predictive
structure, one can reasonably conclude that that method is inadequate
to the task and that one should seek another method.  The nonlinear
LMA method, for instance, performs better in most cases because it is
more general.  The \naive ~method, which simply predicts the mean,
works better in noisy signals because it's a filter.  The dashed line
in Figures blah is a useful empirical guideline for knowing when a
model is not well-matched to the task at hand: predictions whose MASE
scores are more than foo times their WPE are an indication that the
time series has more predictive structure than that method can capture
and exploit.

If information is the limit, then gathering and using more information
is an obvious next step.  There is an equally obvious tension here
between data length and prediction speed: a forecast that requires
half a second to compute is not useful for the purposes of real-time
control of a computer system with a MHz clock rate.  Another
alternative is to sample several system variables simultaneously and
build multivariate delay-coordinate embeddings.  Existing approaches
to that (e.g., \cite{cao-multivariate-embedding,deyle-sugihara2011})
are computationally prohibitive, however.  We are working on
alternative methods that sidestep that complexity.

@@

Nonstationarity is a serious challenge in any time-series modeling
problem.  One of the most common uses of permutation entropy is in
regime-shift detection: e.g., to detect seizures in brainwave data
\cite{cao2004det}.  The SVD signal in Figure~\ref{fig:what} was useful
here because it contains a number of different regimes.  We segmented
this signal visually, but one could imagine using permutation entropy
to do so instead.  This would be a win because it would be automatic,
which would move towards an adaptive modeling strategy, wherein one
would toss the old model at the regime boundary, measure WPE, then use
the result to choose which kind of model to build for the next regime.
More here from Ryan and Joshua.

Experiments here are from one system, but should work on others.  Give
some examples.  Close with mention of what that could enable (viz.,
adaptive models like ones at end of prev paragraph used as the core of
controllers).

>>>>

No forecast strategy will be effective if a time series has no
predictive structure.  In practice, this almost-tautological statement
poses real problems; without knowing anything about the generating
process, it is difficult to determine how much predictive structure is
present in a given time series.  And even if predictive structure
exists, a given forecast method may not work, simply because it cannot
capture and/or exploit the structure that is present (e.g., a linear
model of a nonlinear process).  This too can be a serious practical
challenge.  If a forecast model is not producing good results, a
practitioner needs to know why: is the reason that the time series
contains no predictive structure---i.e., that no model will work---or
is the model that s/he is using simply not good enough?

In this paper, we have argued that redundancy~\cite{crutchfield2003}
is a useful definition of the inherent predictability of a time
series.  To operationalize that definition, we use an approximation of
the Kolmogorov-Sinai entropy~\cite{lind95}, estimated using a weighted
version of the permutation entropy of~\cite{bandt2002per}.  This WPE
technique---an ordinal calculation of forward information transfer in
a time series---is ideal for our purposes because it works with
real-valued data and is known to converge to the true entropy value.
Using a variety of forecast models and more than a hundred time-series
data sets from a computer-performance experiment, we have shown that
prediction accuracy is indeed correlated with weighted permutation
entropy: the higher the WPE, in general, the higher the prediction
error.  There are some exceptions: [[Liz to do: summarize \& discuss
    here.]]

@@

An important practical corollary to this empirical correlation of
predictability and WPE is a practical strategy for assessing
appropriateness of forecast methods.  If the forecast produced by a
particular method is poor but the time series contains a significant
amount of predictive structure, one can reasonably conclude that that
method is inadequate to the task and that one should seek another
method.  The nonlinear LMA method, for instance, performs better in
most cases because it is more general.  The \naive ~method, which
simply predicts the mean, works better in noisy signals because it
effects a filtering operation.  The dashed line in
Figures~\ref{fig:wpe_vs_mase_best} and~\ref{fig:wpe_vs_mase_all} is a
useful empirical guideline for knowing when a model is not
well-matched to the task at hand: a $MASE$ score that is more than
{\color{red}[[what's the equation for that line, Ryan?]]} times the
WPE of the corresponding signal is an indication that that time series
has more predictive structure than that forecast model can capture and
exploit.

Given that information is a fundamental limitation in predictability,
then gathering and using more information is an obvious next step.
But there is an equally obvious tension here between data length and
prediction speed: a forecast that requires half a second to compute is
not useful for the purposes of real-time control of a computer system
with a MHz clock rate.  Another alternative is to sample several
system variables simultaneously and build multivariate models.  This
is a particular challenge in nonlinear LMA-type models, since
multivariate delay-coordinate embedding (e.g.,
\cite{cao-multivariate-embedding,deyle-sugihara2011}) can be
computationally prohibitive.  We are working on alternative methods
that sidestep that complexity.

@@

Nonstationarity is a serious challenge in any time-series modeling
problem.  Indeed, one of the first uses of permutation entropy was
regime-shift detection for the purposes of recognizing seizure onset
in brainwave data~\cite{cao2004det}.  Recall that the signal in
Figure~\ref{fig:svd_ts_colored} was especially useful for the study in
this paper because it contained a number of different regimes.  We
segmented this signal visually, but one could imagine using
permutation entropy to do so instead.  Automating regime-shift
detection would be an important step towards a fully adaptive modeling
strategy, where old models are discarded and new ones are rebuilt
whenever shifts are detected.  WPE would be particularly powerful in
this scenario, as its value can suggest what kind of model might work
well in each new regime.

All of the experiments described in this paper involve a particular
physical system.  We chose that system because its behavior spans a
broad spectrum of classical time series patterns.  We believe that our
results will generalize far beyond this particular system; proving
that claim with examples is another important next step.  [[Liz to do:
    insert some examples.  Close with mention of what that could
    enable (viz., adaptive models like ones at end of prev paragraph
    used as the core of controllers).]]
